Typed Functional Genetic Programming

### 2013-06-29 tex/thesis.tex
[tex] A* and our A* search space ..

@@

\newcommand{\Lets}{Let us }
\newcommand{\lterm}{$\lambda$-term\xspace}
\newcommand{\lterms}{$\lambda$-terms\xspace}
\newcommand{\lhead}{$\lambda$-head\xspace}
\newcommand{\lheads}{$\lambda$-heads\xspace}

@@

\newcommand{\Pseudokod}[2]{
	\begin{framed}
	\begin{algorithm}[H]
		\DontPrintSemicolon
		\SetKwProg{Fn}{Algorithm}{}{}
		\Fn{#1}{#2}
	\end{algorithm}
	\end{framed}
}

@@

@@

\section{Inhabitation tree}

@@

\begin{samepage}
\begin{itemize}
  \item Type nodes   
  			- containing type $\sigma \in \mathbb{T}$ 
  			- aka "OR-node" , Nonterminal-node. 
  \item Symbol nodes 
  			%- containing $ x^{*} \in \mathfrak P \left(V\right) -blobost pač ordered $ or $c \in C$
  			- containing "$\lambda$-head" (nonempty finite sequence of variable names) or constant
  			  name. 
  			- aka "AND-node" , Terminal-node.
\end{itemize}
\end{samepage}

@@

First case \textbf{Atomic type} --- i.e., $\sigma = \alpha$ where $\alpha \in A$:\\
For every $(f,\rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha) \in \Gamma$
where $\alpha \in A$ there is a child \textit{symbol node} of the root containing constant name $f$.
This symbol node containing $f$ has $m$ child subtrees corresponding to Inhabitation trees for 
$(\rho_1,\Gamma),\dots,(\rho_n,\Gamma)$.

@@

By \textit{solving} And-or tree $T$ we mean finding $T'$
subtree of $T$ such that it follows these conditions: 
\begin{itemize}
	\item The root of $T'$ is the root of $T$.
	\item Each \textit{and-node} in $T'$ has all the child nodes as in $T$.
	\item Each \textit{or-node}  in $T'$ has precisely one child 
	      node.\footnote{\textbf{TODO}: MENTION why precisely one 
	      and not at least one ..or CHANGE the def. }   
\end{itemize}
~
\Lets now consider following labeling of Inhabitation tree:

@@

\subsection{Tree representations of \lterms}

@@

\begin{itemize}
	\item $x \in V \cup C$ translates into \textit{leaf} $x$.
	\item $(M N)$ translates into tree\\
		\Tree
			[.@	
		 		\text{$T[M]$}
		 		\text{$T[N]$}		 			
			] 
	\item $\lambda x . M$ translates into tree\\
	 	\Tree
			[.\text{$\lambda x$}	
		 	 	\text{$T[M]$}	
			] 
\end{itemize}

@@

\begin{itemize}
	\item $\lambda x_1 \dots x_n . M$ translates into tree\\
	 	\Tree
			[.\text{$\lambda x_1 \dots x_n$ }	
		 	 	\text{$T[M]$}	
			] 
\end{itemize}

@@

For representing expressions as trees it is however more common use a little different
representation. It will also be the representation suitable for showing 
that \textit{solving} Inhabitation tree generates wanted \lterm.

@@

~\\
\textbf{TODO} 
\begin{itemize}
\item EXAMPLES of tree representations of \lterms  
\item TALK (more?) ABOUT "Barendregt-like" subsection \ref{barlike}\\
		Things about $\beta\eta$ normal form, etc.   
\item DESCRIBE Inhabitation Machine... 
\end{itemize}

\subsection{Our approach to \textit{solving} Inhabitation tree}

@@

\subsubsection{A* algorithm}

@@

As A* traverses the state space, it follows a path of the lowest expected total cost.
It uses a priority queue of alternative unfinished paths.

...\\\\\\\\
\textbf{TODO : čeknout jak je to v kódu jestli ten A* co tam je čekuje ty
 nódy co už navštívil při přidávání
}

\section{Conversion to SKI combinators}

\chapter{ Designed system }	
	\section{Top level view}
		\subsection{Comments about main source files}
			\subsubsection{ Eva.hs }
			\subsubsection{ GP\_{}Core.hs }
	\section{Term generating}
	\section{Crossover}
		\subsection{Finding same types}
		\subsection{Two basic options}
		Resolve problems with free variables or avoid variables completely.

>>>>

% Nejistý fráze který tam všude používám
% ... of a form ...
% Let us ...

@@

\newcommand{\Lets}{Let us\xspace}
\newcommand{\lets}{let us\xspace}
\newcommand{\lterm}{$\lambda$-term\xspace}
\newcommand{\lterms}{$\lambda$-terms\xspace}
\newcommand{\lhead}{$\lambda$-head\xspace}
\newcommand{\lheads}{$\lambda$-heads\xspace}

\newcommand{\la}{\leftarrow\xspace}
\newcommand{\Lp}  {\Lambda^{\prime}\xspace}

@@

\newcommand{\Pseudokod}[2]{
	\begin{framed}
	\begin{algorithm}[H]
		\DontPrintSemicolon
		\SetKwProg{Fn}{function}{}{}
		\Fn{#1}{#2}
	\end{algorithm}
	\end{framed}
}

@@

\section{Conversion to SKI combinators}
\textbf{TODO}

@@

\newpage
\section{Inhabitation tree}

@@

\begin{samepage}
\begin{itemize}
  \item Type nodes   
  			- containing type $\sigma \in \mathbb{T}$ 
  			- aka "OR-node" %, Nonterminal-node. 
  \item Symbol nodes 
  			%- containing $ x^{*} \in \mathfrak P \left(V\right) -blobost pač ordered $ or $c \in C$
  			- containing "$\lambda$-head" (nonempty finite sequence of variable names) or constant
  			  name. 
  			- aka "AND-node" %, Terminal-node. - ale nemusí, proto jsem to zakomentoval
\end{itemize}
\end{samepage}

@@

First case \textbf{Atomic type} --- i.e., $\sigma = \alpha$ where $\alpha \in A$:\\
For every $(f,\rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha) \in \Gamma$
where $\alpha \in A$ there is a child \textit{symbol node} of the root containing constant name $f$.
This symbol node containing $f$ has $m$ child subtrees corresponding to Inhabitation trees for 
$(\rho_1,\Gamma),\dots,(\rho_m,\Gamma)$.

@@

By \textit{solving} And-or tree $T$ we mean finding finite tree $T'$
subgraph of $T$ such that it follows these conditions: 
\begin{itemize}
	\item The root of $T'$ is the root of $T$.
	\item Each \textit{and-node} in $T'$ has all the child nodes as in $T$.
	\item Each \textit{or-node}  in $T'$ has precisely one child 
	      node.\footnote{\textbf{TODO}: MENTION why precisely one 
	      and not at least one ..or CHANGE the def. }   
\end{itemize}
~
\Lets now consider following labeling of Inhabitation tree:

@@

\subsection{Tree representations of \lterms}
\label{tree-reps}

@@

\begin{itemize}
	\item $x \in V \cup C$ translates into \textit{leaf} $x$.
	\item $(P$ $Q)$ translates into tree\\
		\Tree
			[.@	
		 		\text{$T[P]$}
		 		\text{$T[Q]$}		 			
			] 
	\item $\lambda x . P$ translates into tree\\
	 	\Tree
			[.\text{$\lambda x$}	
		 	 	\text{$T[P]$}	
			] 
\end{itemize}

@@

\begin{itemize}
	\item $\lambda x_1 \dots x_n . P$ translates into tree\\
	 	\Tree
			[.\text{$\lambda x_1 \dots x_n$ }	
		 	 	\text{$T[P]$}	
			] 
\end{itemize}

@@

For representing expressions as trees it is however more common to use a little different
representation. It will also be the representation suitable for showing 
that \textit{solving} Inhabitation tree generates wanted \lterm.

@@

~\\
\textbf{TODO} 
\begin{itemize}
\item EXAMPLES of tree representations of \lterms  
\item TALK (more?) ABOUT "Barendregt-like" subsection \ref{barlike}\\
		Things about $\beta\eta^{-1}$ normal form, etc.   
\item DESCRIBE Inhabitation Machine... 
\end{itemize}

\chapter{ A mathematical view of the designed system }

\section{Our approach to \textit{solving} Inhabitation tree}

@@

\subsection{A* algorithm}

@@

As A* traverses the state space, it follows a path of the lowest expected total cost.
It uses a priority queue of alternative unfinished paths.\\

\textbf{TODO : víc to tu rozepsat hlavně zmínit heuristiku 
a že stačí aby byla optimistická atd.. }

\Pseudokod{A*( start , isGoal , heuristic )}{\;

	$open \la$ empty priority queue \;
	$closed \la$ empty set \;\;

	$start$.G = 0 \;
	$start$.F = 0 + $heuristic$( $start$ ) \;\;

	$open$.insert( $start$ )\;\;

	\While{$\neg$ $open$.isEmpty() }{\;

		$state \la$ $open$.popWithLowestF()\;\;

		\If{ isGoal( $state$ ) }{
			\Return $state$ \;
		}\;

		closed.insert( $state$ ) \;\;

		\For{$next \in$ $state$.nexts() }{\;

		    $newG \la$ $state$.G + dist($state$,$next$)\;\;

			\If{ $( next \not\in open \wedge next \not\in closed ) \vee newG < next.$G  }{\;

				$next$.G $\la newG$                          \;
				$next$.F $\la newG + heuristic$( $next$ )  \;\;

				\If{ $next \not\in open$ }{
					$open$.insert( $next$ )\;
				}

			}						
		}
	}\;

	\Return no-path-exists 
}

...\\\\\\\\
\textbf{TODO : čeknout jak je to v kódu jestli ten A* co tam je čekuje ty
 nódy co už navštívil při přidávání - VŽDYŤ díky tomu žeje to tree-search 
 (vs. graph search) to neni potřeba - pořádně se podívat do AIAMA jakj
 je terminologie a napsat to sem.
}

\subsection{Our state space}

% - Stavový prostor 
%   Nedokončený term
%   Navíc unfin typ gama
%   my jen podmnozinu urcenou funkci nexts() a vsemi triv koreny unfin typ gama
%   Vzdy vezmeme nejlevejsi unfin a udelame z nej konkrétnější term pomoci 
%   vyberu v or uzlu tim nam vzniknou 
%   nove unfin uzly odpovidajici pozadavkum and uzlu.
%   Protoze vyplnovani jednotlivych vetvi je na sobe nezavisle 
%   tak je jedno v jakem poradi je budem vyplnovat 
%   a tak je nase "z leva doprava" ok.
% - isFinal(state)
%   Term který neobsahuje unfin uzel
% - Heuristika 
%   simple -pocet unfin uzlů. 
%   Lepsi - predpocitat si to chytre s gama' a udelat si mapu v kery kdyz to neni tak dat 1.
% - Vic nez 1 reseni
%   staci pokracovat pri nalezeni reseni.

In order to describe our state space to search in \lets introduce
extended version of our tree representation for typed \lterms from \ref{tree-reps}.
This extension consists in adding a new type of leaf node $(\sigma,\Gamma)$ 
standing for unfinished tree of a type $\sigma$ with context $\Gamma$ to use
as set of building blocks
\footnote{\textbf{TODO}: Tady to napsat nějak líp s tim gamma }.\\

Initial start states are all of a form "one node tree $(\sigma,\Gamma)$" since we
are trying to generate a typed \lterm of a type $\sigma$ from building blocks $\Gamma$.\\

A tree with no unfinished leaf is considered as final state.  \\

We will define our state space inductively by specifying algorithm for obtaining 
successors of a state.

A non-final state must have one or more unfinished leafs. 
One of those unfinished leafs is selected. 
It is the first one found by depth-first search (from the root of a state tree)
i.e., the leftmost unfinished leaf (if we consider that tree written as expression).

Let $(\sigma,\Gamma)$ be the selected leaf. Successor state is constructed 
by replacing this leaf by new subtree. We must distinguish two cases of $\sigma$ :
\textit{Atomic type} $\sigma$ and \textit{function type} $\sigma$.\\

\textbf{Atomic type} $\sigma = \alpha $, $\alpha \in A$.

For atomic $\sigma$ there are as many successors as there are members of $\Gamma$ of a form 
$(f,\rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \boldsymbol{\alpha} )$,
where $m \geq 0 $. In other words all the members of $\Gamma$ of a type
$\alpha$ or of a type for function which returns $\alpha$.

For each 
$(f,\rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \boldsymbol{\alpha} ) \in \Gamma$,
where $m > 0$ the new subtree which will replace $(\sigma,\Gamma)$ has following form: \\

\Tree
   [.$\alpha$
	[.f	
 		\text{$(\rho_1,\Gamma)$}
 		\text{$(\rho_2,\Gamma)$}
 		\text{$\dots$} 		
 		\text{$(\rho_m,\Gamma)$}		 				 			
	]   
   ]\\

And for m = 0:

\Tree [.$\alpha$ f ] \\

\textbf{Function type} 
$\sigma = \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha$
where $n \geq 1, \alpha \in A$.

For function type $\sigma$ there is exactly one successor.

For every  $i \in \{1,\dots,n\}$ we create new \textit{variable name} $x_i$ which is not yet included in context $\Gamma$ as variable or constant name.

The new subtree which will replace $(\sigma,\Gamma)$ has following form: \\

\Tree
   [.\text{$\tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha$}
	[.\text{$\lambda x_1 \dots x_n$}	
 		\text{$(\alpha,\Gamma \cup \{ (x_1,\tau_1) , \dots , (x_n,\tau_n) \})$}		 				
	]   
   ]\\

\textbf{TODO : Discuss that it is OK to expand only the DFS-found node. 
The order of unfinished nodes is irrelevant since generation of each branch is 
independent on generating other branches. Můžem na to koukat jako na takovou tu redukci 
simetrií, najít to v AIAMA a pořádně popsat.  }

\subsection{Our heuristic function}

An heuristic function takes as input a state and returns as output an estimation of number
of steps needed to get to a final state by the shortest possible path.

Our heuristic function is a simple one: The number of unfinished leafs in the state tree.

\textbf{TODO : rozebrat to, ukázat že je optimistická atd...}

\chapter{ An implementation view of the designed system }

\section{Top level view}
	\subsection{Comments about main source files}
		\subsubsection{ Eva.hs }
		\subsubsection{ GP\_{}Core.hs }
\section{Term generating}
\section{Crossover}
	\subsection{Finding same types}
	\subsection{Two basic options}
	Resolve problems with free variables or avoid variables completely.
